---
title: "Data Science for Public Policy"
subtitle: "Final Project"
author: "Olivia Gomez, Alicia Helfrich, "
execute:
  warning: false
format:
  html:
    embed-resources: true
---

## Final Project

1\. Background and Literature Review: Describes your research question, why the question\
is important to governance, policy, or social sciences, and a brief review of the literature\
that is relevant to your research question. Note that it will likely not be feasible to\
conduct a thorough literature review for this project, but I expect you to show that you\
have at least engaged with the existing research in your focus area.

\
2. Data Sources: This section should describe each of your data sources that you are using\
for your analysis and include your code to access and read in the data. If the data access\
includes any manual steps that are not captured in the code (e.g. downloading files in a\
point-and-click manner) please document those steps here.

```{r}

library(ipumsr)
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(corrplot)
library(VGAM)
library(rpart)
library(rpart.plot)
library(hardhat)
library(ranger)
```

```{r}

# Reading in NHIS data

ddi <- read_ipums_ddi("data/nhis_00002.xml")
nhis <- read_ipums_micro(ddi)
```

\
3. Data Wrangling and Exploratory Data Analysis: This section should include your code\
to perform any data cleaning and new variable creation. You should also thoroughly\
explore your data, including assessing the presence of outliers/unexpected values and\
identifying and appropriately addressing missingness in any key variables.

\
4. Data Analysis: This section should include the code to conduct analysis to answer your\
question of interest. It should include writing explaining why the tools you selected are\
a good fit for the research question and a justification of key analytic decisions. If you\
are using machine learning methods, this section should include model evaluation using\
the methods discussed in class.

\
5. Discussion of Results: This section should include the interpretation of your results.\
Please discuss what your results suggest about the answer to your research question\
of interest. Please also discuss any limitations of your analysis and areas for potential\
future research.
